---
title: "Zonation5R"
output: 
  github_document:
    toc: false
    toc_depth: 2
---

[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)  

*Zonation5R* is an R package that provides an interface to easily integrate and execute Zonation 5 software directly from R. It enables efficient spatial prioritization workflows with enhanced reproducibility and automation. The package was also developed to support teaching and learning, making it easier for R users to get started with Zonation 5 and learn spatial prioritization techniques.

## Installation

You can install the development version from GitHub using:

```{r}
# install.packages("remotes")
# remotes::install_github("thiago-cav/Zonation5R")
```

### Usage

```{r}
library(Zonation5R)
```

### Example 
The **feature list** file is one of the most basic files required to run Zonation. It provides the list of features included in the analysis and their specific parameters. The `create_feature_list()` function generates this from raster files in a specified directory and allows users to include optional attributes based on their requirements.

```{r, eval=FALSE}
# Example usage of the create_feature_list function
create_feature_list(spp_file_dir = "path/to/raster/files",
                    output_dir = "path/to/output",
                    weight = c(0.1, 0.2, 0.3),
                    threshold = 0.5)
```

## Development Notes

This package was developed to facilitate the integration of Zonation 5 with R, supporting spatial prioritization workflows, teaching, and learning. While it is functional and designed to be as useful as possible, I want to clarify that I am not a professional software developer. As such, the development of this package involved many trials and errors, and I’ve greatly benefited from the guidance in [*R Packages (2e)*](https://r-pkgs.org/) by Hadley Wickham and Jennifer Bryan, which provides an excellent foundation for the task. As the authors highlight, *"It doesn’t matter if your first version isn’t perfect as long as the next version is better."* If you encounter any issues or have suggestions for enhancement, feel free to reach out or contribute.


## Author

Thiago Cavalcante  
ORCID: [0000-0001-5357-9659](https://orcid.org/0000-0001-5357-9659)  
Email: thiago.cavalcante@helsinki.fi

