---
title: "Prepare the environment"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Before diving into spatial prioritization with **Zonation 5**, make sure your environment is properly set up and ready to go. Start by:

- Installing the Zonation 5 software on your computer  
- Installing and loading the **ZonationR** package  
- Performing basic checks to ensure your setup is ready for prioritization analyses

This setup is essential before moving on to more advanced workflows and analyses.

### Install Zonation 5 Software

To use **ZonationR**, you must install the **Zonation 5 software** on your computer.

1. Visit the Zonation 5 project site: [zonationteam.github.io](https://zonationteam.github.io/Zonation5/?utm_source=chatgpt.com)

2. Download the appropriate installer or executable for your operating system (e.g., Windows installer, Linux AppImage).

3. Install or unpack the software, and **note the installation path**. You’ll use it later when calling Zonation from R.


### Install the R Package

To install **ZonationR** from GitHub (development version):

```{r, eval=F}
if (!require(devtools)) install.packages("devtools")
devtools::install_github("thiago-cav/ZonationR")
```

Load the package:

```{r, eval=F}
library(ZonationR)
```


### Check Zonation Executable

Before running a full analysis, it’s good practice to verify that R can find the Zonation 5 executable.

```{r, eval=F}
# Check that the executable is valid
check_zonation_executable()
```

If successful, this function will confirm that the path points to the correct Zonation binary.


### Check input layers harmonization

Zonation analyses require that input feature layers (e.g., species distribution rasters) share the same extent, resolution, and alignment. **ZonationR** includes a helper function to **check** whether layers are harmonized.

```{r, eval=F}
# Example feature files (placeholders)

# Check harmonization status of feature layers

```

This function reports if layers differ in extent, resolution, or alignment, helping you identify mismatches before running Zonation and avoiding potential errors.


### What’s Next?

Once installation and checks are done, you’re ready to:

- prepare Zonation input files using `create_feature_list()` and `create_settings_file()`,
- generate an executable call with `create_zonation5_call()`, and
- run a prioritization with `run_zonation5()`.

You can explore these in subsequent vignettes focused on **workflow examples**.


### Further Reading

> If you are new to spatial conservation prioritization or want a gentle overview to understand the basics, see [Why spatial conservation prioritization?](why_spatial_prioritization.html).

Additional resources:

- The **Zonation 5 manual** — detailed instructions and explanations of prioritization settings. ([zonationteam.github.io](https://zonationteam.github.io/Zonation5))



