---
title: "Why spatial conservation prioritization?"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette introduces spatial conservation prioritization, a powerful approach for deciding **where** conservation actions matter most.

## The Geography of Conservation Decisions

Spatial conservation decisions are ultimately decisions about **where to act.**
Even when conservation goals are global, the real impact happens in specific places, whether it's a forest, a coral reef, a river basin, or a broader landscape. Because space is finite, conservation always involves **making choices**.

> Protecting one area usually means not protecting another, at least not immediately or to the same extent.

This simple yet crucial reality shapes modern conservation planning:

- **Biodiversity is unevenly distributed** across the globe.
- **Threats vary spatially**, affecting some areas more than others.
- **Resources are limited**, creating competition among different land uses and priorities.

Therefore, conservation operates under constraints such as:

- Limited budgets,
- Conflicting land uses,
- Multiple objectives that cannot all be maximized simultaneously.

These constraints are **not failures**, they are the **starting point** for effective, strategic conservation planning.

---

## From Spatial Patterns to Decision Challenges

Over the past decades, conservation science has made enormous progress in data availability. Species distribution models, habitat maps, remote sensing products, and global biodiversity databases are now more accessible than ever. However, conservation planners are often challenged not by a lack of data, but by the **difficulty of turning complex, sometimes incomplete data into clear decisions**.

A common first approach is to layer maps on top of one another: visualizing species richness, threat levels, or habitat extent. While informative, this rarely leads to clear choices.

When multiple features are considered simultaneously, **conflicts emerge**:

- Areas important for one species may be less important for another.
- Regions with high biodiversity value may overlap with areas of intense human use.

This often leads to a familiar frustration:

> *“I have all the information I need — so why is deciding still so hard?”*

The problem lies not in the data alone, but in the **lack of a structured way to resolve trade-offs across space**.


## Moving Beyond Yes or No: The Need to Prioritize

Many conservation decisions are often seen as simple **binary choices**: protect or not protect, include or exclude. But in reality, conservation planning rarely works this way. Decisions are usually **incremental** — actions happen over time, budgets change, and priorities evolve as new information becomes available. In this dynamic context, **flexibility is key**.


Instead of asking *“Should this area be protected?”* planners often need to ask:

- **Which areas matter more?**
- **Which actions should happen first?**
- **If only part of a landscape can be addressed now, where should we start?**

This shift from **selection** to **prioritization** fundamentally changes the problem.The goal is no longer to find a single “best” solution, but to support **informed decisions** that work under uncertainty and constraints.

> Understanding this shift helps planners make smarter choices that adapt to real-world challenges.

---



## The need to prioritize, not just select
Many conservation decisions are framed as binary choices: protect or do not protect, include or exclude. In reality, conservation planning rarely works this way. Decisions are typically incremental. Actions are taken over time, budgets change, and priorities are revised as new information becomes available. In this context, flexibility matters. Rather than asking whether an area should be protected at all, planners often need to ask which areas matter more, and which actions should happen first. If only part of a landscape can be addressed in the short term, which locations should receive priority? This shift—from selection to prioritization—changes the nature of the problem. The goal is no longer to define a single optimal solution, but to support informed choices under uncertainty and constraint.


## Ordering space by importance
Spatial conservation prioritization addresses this challenge by ordering locations according to their relative importance for biodiversity conservation. Instead of dividing space into protected and unprotected areas, it produces a ranking that reflects how much each location contributes to overall conservation goals.
This ordering is inherently comparative. The importance of a given place depends on how it complements other places, how many features it supports, and how irreplaceable it is within the broader landscape. A ranked solution allows decision-makers to explore different levels of conservation action, adjust boundaries, and evaluate trade-offs transparently. At this stage, no algorithms or technical details are required to understand the idea. What matters is the conceptual shift: conservation planning becomes a process of ordering space by importance, rather than drawing fixed lines on a map.


## Zonation as a practical implementation
With the problem fully built, Zonation is introduced naturally.
Explain that:
•	Zonation produces a spatial ranking of conservation importance
•	It integrates multiple biodiversity features
•	It supports transparent and reproducible decision-making
At this point, curiosity about how is earned — not forced.


## How this package fits in
This section grounds expectations.
Clarify that ZonationR:
•	does not replace Zonation
•	helps structure and run analyses from R
•	is designed for learning, teaching, and reproducible workflows
This sets up the transition to practical work.

## Installing Zonation
Zonation is external software.
Cover:
•	What installation means in practice
•	Platform differences (Windows vs Linux)
•	Why installation is separate from this package
No commands yet — just orientation.

## Checking your Zonation installation
Introduce validation as a best practice.
Explain:
•	Why checking the executable matters
•	Common configuration issues
Introduce:
•	check_zonation_executable()
Conceptual + light practical framing.

## Validating input feature harmonization
Close with an often-ignored but critical step.
Explain:
•	Why spatial alignment matters
•	Resolution, extent, and projection issues
•	Why errors here propagate silently
Introduce:
•	feature harmonization checks


```{r setup}
library()
```
